---
title: "Data Manipulation"
author: "Ruihan Ding"
date: "2025-09-18"
output: github_document
---

```{r}
library(tidyverse)
```

Import the datasets.

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", ""))
litters_df = 
  janitor::clean_names(litters_df)

pups_df =
  read_csv(
    "data_import_examples/FAS_pups.csv",
    skip = 3,
    na = c("NA", ".", "")
  )
pups_df = 
  janitor::clean_names(pups_df)
```

## `select`

`select` selects variables from my data frame.

select specific variables

```{r}
select(litters_df, group, litter_number, gd0_weight)
```

select a range

```{r}
select(litters_df, group:gd_of_birth)
```

select by removal

```{r}
select(litters_df, -group)
select(litters_df, -(group:gd_of_birth))
```

select by prefix

```{r}
select(litters_df, group, starts_with("gd"))
```

rearrange by selecting

```{r}
select(litters_df, starts_with("gd"), group)
select(litters_df, litter_number, everything())
```

rename by selecting

```{r}
select(litters_df, GROUP = group, everything())
rename(litters_df, GROUP = group)

relocate(litters_df, litter_number)
```

what if I only select one variable?

```{r}
select(litters_df, group)

pull(litters_df, group)

litters_df$group
```

In the pups data, select the columns containing litter number, sex, and PD ears.

```{r}
select(pups_df, litter_number, sex, pd_ears)
```


## `filter`

use `filter()` to filter out rows

```{r}
filter(litters_df, gd_of_birth == 20, pups_born_alive == 6)
```

look for more than two pups alive

```{r}
filter(litters_df, pups_born_alive > 5)
filter(litters_df, pups_born_alive <= 5)
```

look for non-matches

```{r}
filter(litters_df, pups_born_alive != 5) ### != means not equal to
```

match to a set

```{r}
filter(litters_df, group %in% c("Con7", "Con8"))
```

filter out missing rows with `drop_na()`

```{r}
drop_na(litters_df)
drop_na(litters_df, gd0_weight)
```

Learning check 2

```{r}
filter(pups_df, sex == 1)
filter(pups_df, pd_walk < 11, sex == 2)
```


## `mutate`

use `mutate()` to create or modify variables

```{r}
mutate(
  litters_df,
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group)
)
```


## `arrange`

`arrange()` arranges things

```{r}
arrange(litters_df, pups_born_alive)
arrange(litters_df, group, desc(pups_born_alive)) ### desc means descending

arrange(litters_df, gd0_weight)
```


## PIPES

Let's see a _BAD_ way to do things

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", ""))

litters_df1 = 
  janitor::clean_names(litters_df)

litters_df2 = 
  select(litters_df1, starts_with("gd"))

litters_df3 = 
  drop_na(litters_df2)

litters_df4 = 
  mutate(
    litters_df3, 
    wt_gain = gd18_weight - gd0_weight
  )
```

Use pipes to avoid all this

```{r}
litters_df = 
  read_csv("data_import_examples/FAS_litters.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  select(group, litter_number, starts_with("gd")) |>
  drop_na() |>
  mutate(
    wt_gain = gd18_weight - gd0_weight
  )
```

Learning check 3

```{r}
pups_df = 
  read_csv("data_import_examples/FAS_pups.csv", skip = 3, na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  filter(sex == 1) |> ### no need to write 'pups_df' here
  select(-pd_ears) |>
  mutate(pd_pivot_7 = pd_pivot >= 7)
```





